\chapter{Problem Framing, Contributions, and Reading Guide}
\label{ch:framing}

\section{Problem}
This project studies upright-only stabilization of a rotary (Furuta) inverted pendulum on real hardware \cite{Furuta1991,Hamza2019,Ju2025}. The goal is simple: keep the pendulum upright while the rig deals with stepper-actuator limits, sensor noise, missed motion, and real-time timing. The base angle is measured for feedback, but the motor is still driven by open-loop STEP/DIR commands, so the actuator does not behave like an ideal servo.

\section{Key contributions (summary)}
\begin{itemize}
  \item \textbf{Input matched to the firmware command structure.} Most Furuta pendulum treatments use motor torque as the input \cite{Cazzolato2011,Hamza2019}. This project instead uses an acceleration-input model, then implements that input through the firmware path ``acceleration $\rightarrow$ commanded speed $\rightarrow$ stepper''.
  \item \textbf{First-principles modelling of the custom rig.} The report derives the nonlinear model, linearizes it about the upright position, and identifies the constants used later for control design.
  \item \textbf{Three upright controllers on one platform.} The thesis studies one linear full-state controller, one hybrid sliding-mode controller, and one full-surface sliding-mode controller on the same rig.
  \item \textbf{Commanded base positioning while balancing.} A trapezoidal $\theta_{\text{ref}}$ generator with pause/resume logic allows nudge-mode reference moves. In the pinned dataset, this is demonstrated with the linear controller.
  \item \textbf{Practical firmware protections for a stepper-driven rig.} These include derivative filtering, wrap-safe angle differences, glitch rejection, engage logic, soft limits, stall diagnostics, and controlled disarm behavior.
  \item \textbf{A qualitative comparison on a fixed dataset.} Results are drawn from seven pinned runs only: H1--H3, N1, and T1--T3. The comparison is therefore qualitative and metric-based, not statistical.
\end{itemize}

\section{Reading guide}
\begin{itemize}
  \item Hardware and constraints: Chapter~\ref{ch:hardware}
  \item Notation and units: Chapter~\ref{ch:notation}
  \item Modelling derivation (main body): Chapter~\ref{ch:modelling}
  \item Firmware pipeline: Chapters~\ref{ch:measurement}--\ref{ch:actuation}
  \item Controllers: Chapters~\ref{ch:lin}--\ref{ch:smc4}
  \item Experiment protocol and metric definitions: Chapter~\ref{ch:experiments}
  \item Results and comparison: Chapter~\ref{ch:results}
\end{itemize}
