\chapter{Notation, Units, and Recorded Signals}
\label{ch:notation}

\section{Coordinates}
The system is described by two angles:
\begin{itemize}
  \item $\theta(t)$: base or arm yaw angle about the motor axis,
  \item $\alpha(t)$: pendulum angle about the bearing-supported pivot axis, measured from the upright configuration ($\alpha=0$ upright).
\end{itemize}
The state vector used by all controllers is
\[
\mathbf{x}(t)=\begin{bmatrix}\theta & \alpha & \dot{\theta} & \dot{\alpha}\end{bmatrix}^\mathsf{T}.
\]

\section{Sign conventions}
Throughout this report, the sign conventions are chosen such that:
\begin{itemize}
  \item positive $\theta$ corresponds to one fixed arm rotation direction when viewed from above,
  \item positive $\alpha$ corresponds to one fixed pendulum tilt direction relative to the arm.
\end{itemize}
In practice, different builds can differ in wiring and sensor mounting. The sign settings are therefore checked experimentally during bring-up: a small manual pendulum tilt is used to verify the sign of $\alpha$, and a small commanded motor motion is used to verify the signs of $\theta$ and the motor command. For the pinned thesis dataset, the saved settings were \texttt{motorSign = 1}, \texttt{ALPHA\_SIGN = -1}, \texttt{THETA\_SIGN = 1}, and \texttt{CTRL\_SIGN = 1}. The modelling and controller equations in this report assume this sign consistency.

\section{Wrap-safe angle differences}
Both encoders provide absolute angles in $[0,360^\circ)$. For control and reference tracking we use the shortest signed angle difference:
\[
\mathrm{diff}(a,b)\in[-180^\circ,180^\circ],\qquad \mathrm{diff}(a,b)\equiv a-b\ \text{wrapped to}\ [-180^\circ,180^\circ].
\]
This avoids discontinuities at $0/360^\circ$. It is used, for example, for the base tracking error
\[
\theta_{\mathrm{err}}=\mathrm{diff}(\theta,\theta_{\mathrm{ref}}).
\]
In firmware this means \emph{current minus reference}, matching the helper \texttt{getAngleDiffDeg}.

\section{Units policy and stepper conversion}
The first-principles modelling in Chapter~\ref{ch:modelling} is expressed in SI units (radians, seconds). The embedded implementation and plots in this report use degrees for angles for readability.

The stepper driver is commanded in \si{\step\per\second}. A constant conversion between angle and steps is used:
\[
k_s \equiv \frac{\text{steps}}{\text{degree}},
\]
so that $\theta$ can be converted to measured steps by $k_s\,\theta$. Likewise, an angular acceleration command (\si{\degree\per\second\squared}) can be converted to a stepper acceleration command (\si{\step\per\second\squared}) by multiplying by $k_s$.

When this report writes $u=\ddot{\theta}$ in the model chapters, $u$ is the abstract acceleration input. In firmware and plots, the logged command is the implementation-space version of that input, expressed in stepper units.

\section{Variables and recorded signals}
This report uses three related but different naming layers:
\begin{itemize}
  \item \textbf{Model variables:} $\theta$, $\alpha$, $\dot{\theta}$, $\dot{\alpha}$, and $u=\ddot{\theta}$.
  \item \textbf{Firmware variables:} names used directly in the embedded code.
  \item \textbf{Raw CSV columns:} the columns written directly by the logger.
\end{itemize}

\subsection*{Model variables}
\begin{itemize}
  \item $\theta$, $\alpha$: measured base and pendulum angles,
  \item $\dot{\theta}$, $\dot{\alpha}$: estimated angular rates,
  \item $u=\ddot{\theta}$: the control input used in the model and control derivations.
\end{itemize}

\subsection*{Firmware variables}
\begin{itemize}
  \item \texttt{thetaErr = diff(theta, thetaRef)}: base tracking error,
  \item \texttt{thetaDotErr = thetaDot - thetaRefDot}: base-rate tracking error,
  \item \texttt{accCmdPhysical}: commanded base acceleration before motor-sign mapping,
  \item \texttt{thetaDotCmdMotor}: commanded motor speed after acceleration integration.
\end{itemize}

\subsection*{Raw CSV columns}
Each experimental trial records:
\begin{itemize}
  \item \texttt{alpha\_deg}, \texttt{alpha\_dot\_deg\_s},
  \item \texttt{theta\_deg}, \texttt{theta\_dot\_deg\_s},
  \item \texttt{acc\_cmd\_steps\_s2}, \texttt{vel\_cmd\_steps\_s},
  \item \texttt{pos\_meas\_steps}, \texttt{position\_clamped}.
\end{itemize}
The raw CSV does \emph{not} contain a dedicated acceleration-saturation column. Saturation percentages are computed later by the analysis pipeline from the logged command and the known firmware limits.

In addition to the high-rate CSV stream, the firmware also prints low-rate status and diagnostic lines used to segment trials and, when possible, align host-time events to device-time signals (Chapter~\ref{ch:experiments}).
