\chapter{Linear Full-State Feedback Controller}
\label{ch:lin}

\section{Linearized plant for control design}
This controller is the baseline controller used in the thesis. It uses all four states and tries to keep the pendulum upright while also keeping the arm near the reference angle.

The controller is designed about the upright equilibrium ($\alpha \approx 0$) using the acceleration-input formulation $u \equiv \ddot{\theta}$ from Chapter~\ref{ch:actuation}. Using the linearized model derived in Chapter~\ref{ch:modelling}, the upright dynamics can be written as
\[
\mathbf{x} \;=\; \begin{bmatrix}\theta & \alpha & \dot{\theta} & \dot{\alpha}\end{bmatrix}^{\mathsf{T}}
\;\equiv\; \begin{bmatrix}x_1 & x_2 & x_3 & x_4\end{bmatrix}^{\mathsf{T}}
\]
and
\begin{align}
\dot{x}_1 &= x_3, &
\dot{x}_2 &= x_4, &
\dot{x}_3 &= u, &
\dot{x}_4 &= A\,x_2 - B\,u.
\label{eq:lin_plant}
\end{align}
For this rig,
\[
A = 100.8,\qquad B = 1.952.
\]
In state-space form,
\[
\dot{\mathbf{x}} = \mathbf{A}\mathbf{x} + \mathbf{B}u,\qquad
\mathbf{A}=
\begin{bmatrix}
0&0&1&0\\
0&0&0&1\\
0&0&0&0\\
0&A&0&0
\end{bmatrix},\qquad
\mathbf{B}=
\begin{bmatrix}
0\\0\\1\\-B
\end{bmatrix}.
\]
Linear state feedback is a common baseline for rotary inverted pendulum stabilization \cite{Park2011,Chawla2018}.

\section{Full-state feedback by pole placement}
Choose the state-feedback law
\begin{equation}
u = -\mathbf{K}\mathbf{x},\qquad
\mathbf{K}=\begin{bmatrix}k_\theta & k_\alpha & k_{\dot{\theta}} & k_{\dot{\alpha}}\end{bmatrix}.
\label{eq:lin_state_feedback}
\end{equation}
Substituting \eqref{eq:lin_state_feedback} into \eqref{eq:lin_plant} gives the closed-loop matrix
\[
\mathbf{A}_{\mathrm{cl}}=\mathbf{A}-\mathbf{B}\mathbf{K}
=
\begin{bmatrix}
0 & 0 & 1 & 0\\
0 & 0 & 0 & 1\\
-k_\theta & -k_\alpha & -k_{\dot{\theta}} & -k_{\dot{\alpha}}\\
B k_\theta & A + B k_\alpha & B k_{\dot{\theta}} & B k_{\dot{\alpha}}
\end{bmatrix}.
\]
Eliminating $\theta$ from the scalar equations gives the fourth-order characteristic polynomial
\begin{equation}
s^4 + a_3 s^3 + a_2 s^2 + a_1 s + a_0 = 0,
\label{eq:lin_charpoly}
\end{equation}
with coefficients
\begin{align}
a_3 &= k_{\dot{\theta}} - B k_{\dot{\alpha}}, &
a_2 &= k_{\theta} - A - B k_{\alpha}, &
a_1 &= -A k_{\dot{\theta}}, &
a_0 &= -A k_{\theta}.
\label{eq:lin_coeffs}
\end{align}
This form is useful because the gains can be found directly by matching a desired polynomial \cite{ogata_modern_control}.

\subsection{Desired poles via two second-order factors}
The desired closed-loop polynomial is chosen as a product of two second-order factors:
\begin{equation}
\bigl(s^2+2\zeta_1\omega_1 s+\omega_1^2\bigr)\,
\bigl(s^2+2\zeta_2\omega_2 s+\omega_2^2\bigr)
= s^4 + \tilde a_3 s^3 + \tilde a_2 s^2 + \tilde a_1 s + \tilde a_0,
\label{eq:lin_desired_poly}
\end{equation}
where $(\omega_1,\zeta_1)$ describe a faster pendulum mode and $(\omega_2,\zeta_2)$ describe a slower base-centering mode. Expanding gives
\begin{align}
\tilde a_3 &= 2(\zeta_1\omega_1+\zeta_2\omega_2),\nonumber\\
\tilde a_2 &= \omega_1^2+\omega_2^2+4\zeta_1\zeta_2\omega_1\omega_2,\nonumber\\
\tilde a_1 &= 2(\zeta_1\omega_1\omega_2^2+\zeta_2\omega_2\omega_1^2),\nonumber\\
\tilde a_0 &= \omega_1^2\omega_2^2.
\label{eq:lin_desired_coeffs}
\end{align}

\subsection{Closed-form gains}
Setting $(a_0,a_1,a_2,a_3)=(\tilde a_0,\tilde a_1,\tilde a_2,\tilde a_3)$ and solving \eqref{eq:lin_coeffs} yields
\begin{align}
k_{\theta} &= -\frac{\tilde a_0}{A}, &
k_{\dot{\theta}} &= -\frac{\tilde a_1}{A},\nonumber\\
k_{\dot{\alpha}} &= \frac{k_{\dot{\theta}}-\tilde a_3}{B}, &
k_{\alpha} &= \frac{k_{\theta}-A-\tilde a_2}{B}.
\label{eq:lin_gain_formulas}
\end{align}

\subsection{Numerical example and unit conversion}
Using $A=100.8$, $B=1.952$ and choosing a fast mode $(\omega_1,\zeta_1)=(15,0.8)$ and a slow mode $(\omega_2,\zeta_2)=(1,1)$, the desired coefficients are
\[
\tilde a_3=26,\qquad \tilde a_2=274,\qquad \tilde a_1=474,\qquad \tilde a_0=225,
\]
and the resulting gains from \eqref{eq:lin_gain_formulas} are
\[
\mathbf{K}=\begin{bmatrix}
-2.232 & -193.15 & -4.702 & -15.729
\end{bmatrix}.
\]
For implementation with degree states and stepper units, define the conversion $k_s$ (units \si{\step\per\degree}; Chapter~\ref{ch:notation}). The commanded acceleration in stepper units is
\[
u_{\mathrm{steps}} = k_s\,u_{\deg}.
\]
Therefore, the gain vector expressed in stepper-acceleration units is
\[
\mathbf{K}_{\mathrm{steps}} = k_s\,\mathbf{K}.
\]
With $k_s=4.444$, the example yields
\[
\mathbf{K}_{\mathrm{steps}}=
\begin{bmatrix}
-9.92 & -858.4 & -20.9 & -69.9
\end{bmatrix}.
\]
This is a \emph{design example}, not the final tuned parameter set used in the experiments.

\section{Reference tracking and nudge behavior}
Rather than regulating the base to $\theta=0$ all the time, the implemented controller uses reference tracking:
\[
\theta_{\mathrm{err}} = \mathrm{diff}(\theta,\theta_{\mathrm{ref}}),\qquad
\dot{\theta}_{\mathrm{err}} = \dot{\theta} - \dot{\theta}_{\mathrm{ref}},
\]
with $\mathrm{diff}(\cdot,\cdot)$ the wrap-safe difference operator from Chapter~\ref{ch:notation}. The reference-tracking full-state feedback law is
\begin{equation}
u_{\mathrm{cmd}} =
u_{\mathrm{ff}} +
K_\theta\,\theta_{\mathrm{err}} + K_\alpha\,\alpha +
K_{\dot{\theta}}\,\dot{\theta}_{\mathrm{err}} + K_{\dot{\alpha}}\,\dot{\alpha},
\label{eq:lin_ref_tracking}
\end{equation}
with feedforward term $u_{\mathrm{ff}}=\ddot{\theta}_{\mathrm{ref}}$ provided by the reference generator.

\subsection*{Commanded base position while balancing (nudge mode)}
The reference $\theta_{\mathrm{ref}}$ is generated by a velocity- and acceleration-limited trapezoidal profile driven by a user-specified base angle target. This gives practical commanded base positioning while keeping the pendulum upright.

Because aggressive base motion can destabilize the pendulum, the reference generator includes a pause/resume mechanism. When instability is confirmed, the reference is handed once to the current measured $\theta$ and then held there until the pendulum becomes stable again. Once the pendulum returns to the stability window, the profile resumes.

In implementation, the move stability gate is defined as $|\alpha|<\alpha_{\mathrm{move,win}}$ with $\alpha_{\mathrm{move,win}}=\SI{5}{\degree}$ and $|\dot{\alpha}|<\dot{\alpha}_{\mathrm{move,win}}$ with $\dot{\alpha}_{\mathrm{move,win}}=\SI{80}{\degree\per\second}$. The gate is enabled only after a post-engage grace $T_g=\SI{200}{ms}$. Stability and instability are each debounced for $10$ consecutive control ticks (\SI{50}{ms} at \SI{200}{Hz}) before pausing or resuming the profile. For safety, user targets are also bounded by $|\theta_{\mathrm{target}}|\le \SI{35}{\degree}$.

\section{Implementation details (stepper-oriented)}
Several practical additions are used to improve repeatability on real hardware:
\begin{itemize}
  \item \textbf{Deadband on $\alpha$:} $\alpha_{\mathrm{db}}=\SI{0.5}{\degree}$ reduces noise-chasing near upright.
  \item \textbf{Acceleration saturation and engage ramp:} the commanded acceleration is saturated to $u_{\max}=\SI{20000}{\step\per\second\squared}$, the commanded speed is capped at $\dot{\theta}_{\max}=\SI{15000}{\step\per\second}$, and the command is multiplied by a linear engage ramp of duration $T_{\mathrm{ramp}}=\SI{100}{ms}$ after controller engagement.
  \item \textbf{Low-speed stop deadband:} the motor is stopped for very small speed commands using \texttt{speedStopHz} set to \SI{50}{\step\per\second}.
  \item \textbf{Velocity leak (linear-only):} $\lambda_{\mathrm{leak}}=\SI{2.0}{s^{-1}}$ is enabled only when $|\alpha|<\alpha_{\mathrm{leak,win}}=\SI{5}{\degree}$ and $|\theta_{\mathrm{err}}|<\theta_{\mathrm{leak,win}}=\SI{20}{\degree}$ to suppress slow drift.
  \item \textbf{Auto-trim of the upright reference:} enabled only when $|\alpha|<\alpha_{\mathrm{trim,win}}=\SI{3}{\degree}$ and $|\dot{\alpha}|<\dot{\alpha}_{\mathrm{trim,win}}=\SI{50}{\degree\per\second}$; integrates at $\lambda_{\mathrm{trim}}=\SI{0.20}{s^{-1}}$ and is clamped to $|\alpha_{\mathrm{trim}}|\le \SI{1.5}{\degree}$ relative to the calibration reference.
  \item \textbf{Soft limits:} outward velocity commands are suppressed within a margin $\Delta_\theta=\SI{2}{\degree}$ of the mechanical base limit.
\end{itemize}

\section{Parameters used in experiments}
The tuned firmware gains used in the pinned dataset are not the same as the pole-placement example above. The deployed gains differ because the implementation separates sign conventions from gain magnitudes, uses stronger base centering, and includes practical hardware tuning on the real rig.

\begin{table}[htbp]
  \centering
  \small
  \setlength{\tabcolsep}{5pt}
  \caption{Linear controller gains used in the pinned experiments (stepper acceleration units).}
  \label{tab:lin_gains}
  \begin{tabular}{lrrl}
    \toprule
    Parameter & Value & Units & Meaning \\
    \midrule
    $K_{\theta}$ & 31.5 & \si{\step\per\second\squared\per\degree} & base or arm position gain \\
    $K_{\alpha}$ & 725.6 & \si{\step\per\second\squared\per\degree} & pendulum angle gain \\
    $K_{\dot{\theta}}$ & 28.3 & \si{\step\per\second\squared}\,per\,(\si{\degree\per\second}) & base or arm rate gain \\
    $K_{\dot{\alpha}}$ & 73.5 & \si{\step\per\second\squared}\,per\,(\si{\degree\per\second}) & pendulum rate gain \\
    \bottomrule
  \end{tabular}
\end{table}
Derivative estimation uses the filtered differentiator described in Chapter~\ref{ch:measurement} with cutoff $\omega_c=\SI{450}{rad/s}$.

\ObservedBox{See Figures~\ref{fig:hold_compare_timeseries}, \ref{fig:nudge_lin_tracking}, \ref{fig:tap_compare_timeseries} and Tables in Chapter~\ref{ch:results}.}
